---
title: "class 9 structural bioinformatics (pt 1)"
author: "Trinity Leahy"
format: pdf
editor: visual
---

# PDB Statistics

I am going to download a CSV file from the PDB site and move it into my R studio.

```{r}
db <- read.csv("Data Export Summary.csv")
db
```

> Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.

First I created a function so that I can convert my string of numbers with a comma for each column into characters without a comma that R can read as a numeric value and then add them together.

```{r}
# I will work with `x` as input.

convertandsum <- function(x){
  #Substitute the comma and convert it into a number
  sum(as.numeric(gsub(",", "", x)))
}
```

For Xray:

```{r}
round(convertandsum(db$X.ray)/convertandsum(db$Total), 2)
```

For Electron Microscopy:

```{r}
round(convertandsum(db$EM)/convertandsum(db$Total), 2)
```

> Q2: What proportion of structures in the PDB are protein?

```{r}
round(convertandsum(db$Total[1])/convertandsum(db$Total), 2)
```

> Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

Skipped!

> Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

The structure is too low resolution to see H atoms. You need a sub 1 angstrom resolution to see hydrogen.

> Q5: There is a critical “conserved” water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have

Yes, it is HOH308.

> Q6: Generate and save a figure clearly showing the two distinct chains of HIV-protease along with the ligand. You might also consider showing the catalytic residues ASP 25 in each chain and the critical water (we recommend “Ball & Stick” for these side-chains). Add this figure to your Quarto document.

![](1HSG (2).png)

# Working with Structures in R

We can use the `bio3d` package to read and perform bioinformatics calculations on PDB structures.

```{r}
library(bio3d)

pdb <- read.pdb("1HSG")
pdb
```

```{r}
attributes(pdb)
```

> Q7: How many amino acid residues are there in this pdb object? 

  198 amino acid residues

> Q8: Name one of the two non-protein residues? 

  MK1
  
> Q9: How many protein chains are in this structure? 

  2 chains

```{r}
head(pdb$atom)
```

Read an ADK structure

```{r}
adk <- read.pdb("6s36")
adk
```

Perform a prediction of flexibility with a technique called NMA (normal mode analysis).

```{r}
# Perform flexiblity prediction
m <- nma(adk)
plot(m)
```

Write out a "movie" (a.k.a. trajectory) of the motion for viewing in Molstar.

```{r}
mktrj(m, file="adk_m7.pdb")
```

