---
title: "Class 05: Data Visualization"
author: "Trinity (PID: A15766955)"
format: pdf
editor: visual
---

# Base R graphics vs ggplot2

There are many graphics systems available in R, including so-called "base" R graphics and the very popular **ggplot2** package.

To compare there let's play with the inbuilt `cars` dataset.

```{r}
head(cars)
```
To use "base" R I can simply call the `plot()` function: 
```{r}
plot(cars)
```

To use `ggplot2` package, I first need to install it with the function`install.packages("ggplot2")`.

I will run this in my R console (i.e. the R brain) as I do not want to re-install every time I render my report...

```{r}
library(ggplot2)
ggplot()
```

To make a figure with ggplot, I always need three things:

 - **data** (what I want to plot)
 - **aes** (the aesthetic mapping of the data)
 - **geoms** (how I want to plot the data)

```{r}
ggplot(data=cars) + 
         aes(x=speed, y=dist) +
         geom_point()
```
I can just keep adding layers.

```{r}
ggplot(data=cars) + 
         aes(x=speed, y=dist) +
         geom_point() + geom_smooth()
```

Ggplot is much more verbose than base R plots for standard plots but it has a consistent later system that I can use to make just any plot.

Let's make a plot with a straight line fit - i.e. a linear model and no standard error shown.

```{r}
ggplot(data=cars) + 
         aes(x=speed, y=dist) +
         geom_point() + geom_smooth(se = FALSE, method = lm)
```

```{r}
ggplot(data=cars) + 
 aes(x=speed, y=dist) +
 geom_point() + 
 labs(title="speed vs stopping distance of cars", x ="speed (MPH)", y="stopping distance (ft)", subtitle="BIMM 143", caption="Dataset: 'cars'") +
 geom_smooth(se = FALSE, method = "lm") + 
 theme_bw()
```

## A more complicated plot

Let's plot some gene expression data

Let’s turn for a moment to more relevant example data set. The code below reads the results of a differential expression analysis where a new anti-viral drug is being tested.
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```

> Q. How many columns are there and what are their names?

```{r}
colnames(genes)
ncol(genes)
```

>Q. How can we summarize that last column - the "state" column?

```{r}
table(genes$State)
```

Plotting this data:

```{r}
p <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, color=State) + 
  geom_point()
```

```{r}
p + labs(x="control", y="drug treatment")+ scale_colour_manual(values=c("green", "beige", "blue"))
```

## Going Further

Here I need a slightly larger dataset.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.3)
```

A very useful layer to add sometimes is "faceting"

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.3) +
  facet_wrap(~continent)
```
