>Q6. How would you generalize the original code above to work with any set of input
protein structures?

This is the original code:

```{r}
# Can you improve this analysis code?
library(bio3d)

s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

I will begin by calling package `bio3d` up from my library.
```{r}
library(bio3d)
```

#The Breakdown

Next, I am going to break down each code structure to determine what they do and if they are functional. 

```{r}
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
```
The above lines of code use the unique PDB accession codes in order to access a file about a specific protein.

```{r}
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
```

The `trim.pdb()` function calls on R to access a specific chain within the protein we are interested in. The input arguments are (protein (defined by pdb), chain (defined by what chain we are interested in), and electy (character vector of atom names)).

```{r}
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
```

This code even further specifies the part of the protein we are looking at.

```{r}
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

This code plots the information we are looking at pertaining to our protein.

# Simplifying

Now I am going to simplify the code so that it is more straight forward and concise.

```{r}
#calling forward the package bio3d
library(bio3d)

#specifying what protein and which aspect it we want to examine
s1.chainA <- trim.pdb(read.pdb("4AKE"), chain="A", elety="CA")

#plotting the data for the specific part of the protein
plotb3(s1.chainA$atom$b, sse=s1.chainA, typ="l", ylab="Bfactor")
```

# Function Creation

Finally, I am going to transfer my more concise code into a function.

```{r}
examine.protein <- function(x){
  
  #calling forward the package bio3d
  library(bio3d)

  #specifying what protein and which aspect it we want to examine
  chainA <- trim.pdb(read.pdb(x), chain="A", elety="CA")

  #plotting the data for the specific part of the protein
  plotb3(chainA$atom$b, sse=chainA, typ="l", ylab="Bfactor")}
```

And now we check to make sure that the function does what we are looking for!

```{r}
examine.protein("4AKE")
examine.protein("1AKE")
examine.protein("1E4Y")
```

The `examine.protein()` function I created takes the **input** of a PDB code and specifies the part of the protein we are looking at to be a specific part of chain A and then **outputs** a plot of the B-Factor vs Residue data for that part of the protein. It can be used for any protein with a PDB code and multiple chains to look at the chain A of the protein.