---
title: "Class 17: Investigating Pertussis"
author: "Trinity Leahy"
format: pdf
editor: visual
---

Pertussis, or whooping cough, is a highly contagious lung infection caused by a bacteria *B. pertussis*.

The CDC tracks reported cases in the U.S. since the 1920s.

#Investigating Pertussis Cases by Year

```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

```{r}
library(ggplot2)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
baseplot <- ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Percussis Cases by Year (1922-2021)") +
  xlab("year") +
  ylab("Number of Cases")
baseplot
```

#The Tale of Two Vaccines

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

The first big "whole-cell" pertussis vaccine program started in 1942.

```{r}
baseplot +
  geom_vline(xintercept = 1942, color = "light blue") +
  geom_vline(xintercept = 1980, color = "grey", linetype = 2) +
  geom_vline(xintercept = 1996, color = "dark blue")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, we see a very gradual uptick in the amount of cases. There is a couple of possible explanations for this trend: some people could still be vaccine hesitant, *the new vaccine could be less effective long-term*, or the bacteria could be evolving.

# Exploring CMI-PB Data

Something big is happening with pertussis cases and big outbreaks are once again a major public health concern! BUGGER

One of the main hypothesis for the increasing case numbers is waning vaccine efficiency with the newer aP vaccine.

Enter the CMI-PB project, which is studying this problem on large scale. Let's see what data they have.

Their data is available in JSON format ("key:value" pair style)
We will use the "jsonlight" package to read their data.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
```

```{r}
head(subject, 3)
```

> Q4. How may aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)

head(specimen, 6)
```

I want to "merge" the `subject` and `specimen` tables together. I will use the **dplyr** package for this.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```


```{r}
ncol(specimen)
ncol(subject)
ncol(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
ab <- read_json("http://cmi-pb.org/api/ab_titer", simplifyVector=TRUE)
head(ab)
```

```{r}
abdata <- inner_join(meta, ab)
head(abdata)
```

```{r}
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```
There are way less visit 8 specimens because the project is still ongoing. The 8th visits are happening now and therefore we do not have that data for all individuals yet.

# Examine IgG1 Ab Titer Levels

We will use the `filter()` function from dplyr to focus on just IgG1 isotype and visits 1 to 7.

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow = 1)
```

> Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

FIM2/3 (Fimbrial Protein) clearly changes. This is a protein involved in making bacteria pilus and is involved in cell adhesion
FHA (Filamentous hemagglutinin) is secreted onto the cell surface of bacteria.
DT (Diphtheria Toxin) 
PT, PTM, PRN

All the above antigens are involved in extracellular function that should rise and change when exposed to a vaccine.

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).

```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time course and the FIM2/3 data in particular?

The median MFI for the measles antigen is consistent throughout time, regardless of the type of vaccine the person was exposed to. This is expected as the measles antigen is a control. However, the FIM2/3 vaccine appears to be a lot more variable across the visit days as well as when comparing which vaccines the people had been exposed to.

> Q17. Do you see any clear difference in aP vs. wP responses?

There is possibly a clear difference in aP vs wP response for day 7 visit for the FIM2/3 but more statistical analysis is needed to know for sure.



