---
title: "class10"
author: "Trinity Leahy"
format: html
editor: visual
---

#Comparative analysis of ADK

ADK (Adenelate Kinase) is an important drug target and we would love to know how it works - i.e. molecular structure.

```{r}
install.packages("bio3d")
install.packages("devtools")
install.packages("BiocManager")

BiocManager::install("msa")
```

```{r}
devtools::install_bitbucket("Grantlab/bio3d-view")
```

> Q10. Which of the packages above is found only on BioConductor and not CRAN? 

msa package

> Q11. Which of the above packages is not found on BioConductor or CRAN?: 

Grantlab/bio3d-view package

> Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket? 

TRUE

```{r}
library(bio3d)
aa <- get.seq("1ake_A")
aa
```

> Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 

214 AA

We can nbow run BLAST with this sequence.

```{r}
b <- blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```


```{r}
head(hits$pdb.id)
```

Now we can download all these PDB files.

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Now I want to align and superimpose these structures which are all over the place.

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile = "msa")
```

Let's have a look at our `pdbs` objects.

```{r}
pdbs
```

Now we have our aligned and superimposed structures we can perform all sorts of analysis on them. Let's do PCA...

```{r}
pc.xray <- pca(pdbs)
plot(pc.xray)
```

Results of PCA on Adenylate kinase X-ray structures. Each dot represents PDB structure.

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

We can cluster the structures by RMSD (or any other method).

We can make a wee movie - also called a trajectory of the major differences (i.e. structural displacements) of ADK.

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```


